Asymmetric tail-dependence in extreme climatic events affects population-extinction risk

### 2019-08-29 MainText.Rmd
intro edited

With global climate change, extreme climatic events (ECEs) appear to be more common than usual. These ECEs 
could pose severe impact on the resistance or resilience of a population or community level through
life-history traits of the species. Therefore predicting the impact of ECEs 
on extinction risk of a population is necessery. We here introduced the potential use of "tail-dependence" 
property of a copula in modelling the extinction risk of an age-structured single patch model. We also 
extended the generalized findings of density-independent metapopulation model from @Ghosh_copula 
to density-dependence metapopulation models with this copula approach. For metapopulation models, 
we showed extinction risk not only depended on the nature of tail-dependence of spatially correlated 
noise but was also influenced by the inherent dynamics of the population-model.

@@

<!--Earlier study on metapopulation extinction risk-->

<!--How copula could be relevent in estimating extreme event's 
influence on extinction risk : introduce Taildep. concept-->
A recent study [@Ghosh_copula] used the tail dependence feature 
to estimate extinction risk of a density-independent 
metapopulation model. Present manuscript is an extension for the application of copula 
in estimating the extinction risk as introduced by [@Ghosh_copula]. We will show 
why the potential use of copula in modelling extinction risk goes beyond the density-independent model 
and even applicable for a simple age-structured population where its vital rates are influenced by ECEs. 
Following we will introduce the concept of copula in a manner that follows @Ghosh_copula.

<!--brief intro for copula-->
Copula, a rank based measure of association, is much more informative 
than linear correlation measure and a widely applied tool in dependence 
modeling [@joe2014_dependence; @nelsen2006_copula]. Copulas separate the information content of a bivariate dataset,
$(x_n, y_n)$ for $n = 1 \ldots N$, into two non-overlapping parts: the information in the marginal 
distributions (which is not about the association between the variables) and the rest of 
the information (which is solely about the association). The association information is revealed by 
the copula plot ($v_n$ against $u_n$) of the raw variables ($y_n$ vs. $x_n$) in Fig...., where 
$u_n$ is the rank of $x_n$ in the set ${x_1, x_2, \ldots , x_n }$, divided by $N + 1$; 
and $v_n$ is the rank of $y_n$ in the set ${y_1, y_2, \ldots , y_n }$, also divided
by $N + 1$. Here the rank of the smallest element of each sets is considered to be 1 and the rank of
the largest element is $N$. We refer to the $u_n$ and $v_n$ as normalized ranks of the $x_n$ and $y_n$. 
Ranking makes the marginal distributions uniform, isolating only the information on 
association between the variables.

<!--tail dep. of copula & extinction risk-->
A major benefit of the copula approach is that it can effectively capture the tail dependence. 
Tail dependence, already exemplified in Fig..... (add Fig.), is association between the more extreme 
values of distributions. For two positively associated variables, left-tail 
(respectively, right-tail) dependence is stronger association between
values in the left or lower portions, i. e. for smaller values 
(respectively, the right or upper portions, i.e. for higher values) 
of the two distributions. Stronger dependence in either tail gives rise to asymmetric copula 
structure for the joint distribution of two variables. 
For example, extreme cold winter during growing season
might reduce the adult reproduction rate as well as larval survival (i.e. stronger 
tail dependence between vital rates when temperature is too cold than normal), which 
could lead to population crash and thereby inflate the extinction risk. Similarly, in context of 
density-dependent metapopulation models, if extreme environmental fluctuations across sites 
appear to be as synchronous (left tail dependent or right tail dependent between patches 
with same correlation as in Fig.....) then it could wipe out some metapopulation patches 
during unfavorable years unless they are rescued by the nearby populations. This scenario 
could be more complex depending 
on the dynamics of the model (see \nameref{Results}). Though tail-dependence 
property of a copula has been used for risk-analysis in other fields like finance, 
hydrology etc. [@alexander2008market; @li2013; @chen2011flood; @Goswami2018], but to our knowledge
copula have never been used in estimating the effect of ECEs through extinction risk of age structured 
model or density-dependent metapopulations.

<!--our objective ?-->
In this paper, we will explore the potential use of asymmetric tail-dependence in a copula 
for understanding the extinction risk of populations towards extreme climatic event
by addressing the following questions: (Q1) Does nature of tail dependence 
(left tail or right tail) in vital rates
of a single-patch age-structured population model affect extinction risk of that population differently?
We hypothesized population might have higher extinction risk if vital rates are left-tail dependent.
(Q2) How does environmental noises with asymmetric tail dependence across multiple patches
influence extinction risk of a density-dependent metapopulation model? Our speculation 
is that the general finding from density-independent metapopulation model [@Ghosh_copula]
(higher extinction risk if noises across patches were left-tail dependent than the one 
if noises had right-tail dependence) would hold true only in the under-compensatory regime, but 
should be reverse if the model shows over-compensatory dynamics. 
(Q3) We further hypothesize that when there is no dispersal, the above-said switching should occur
near the critical parameter value where the deterministic model shows transition from 
under-compensatory to over-compensatory dynamics.

@@

<!--future study: worthy for PVA, IUCN red list
1. Identify which life stage or demographic process as management target?
2. Old study (Doak1994, Ferguson1995) showed including correlation btw vital rates for PVA makes a difference, 
   what if we include TD through copula? Multisite age str. model with dispersal, directional change
   in spatio-temporal variation for vital rates.
3. Tdep. with -ve correlation between vital rates?
4. extinction risk how propagated through a food web?
-->
Although we demonstrated that tail dependence between vital rates of an age structured model can
influence extinction risk, substantial work remains to determine the importance of this effect in the 
following scenarios. First, we demonstrated the effect of tail-dependence on extinction risk 
with a very simplistic age structured (two age stages) model where we considered the total semelperous 
population would go extinct if there would be only one female adult and one egg. But in reality population 
viability analysis (PVA) deals with a little bit higher threshold with several age, stage or size class and a variety 
of populations including semelperous and iteroperous [@morris2002quantitative]. Future studies should include 
tail-dependence effect on extinction risk 
for a multiple age-, size- and stage- structured model and estimate extinction risk 
for each age, size and stage class separately 
so that we can identify (via multisite-PVA) which life-stage or demographic process 
to target for population management. Earlier study by @doak1994modeling showed PVA models including correlation 
between different demographic rates indicated marked difference in predicting confidence interval for 
future population size than the model which did not include correlation. But when 
such correlation cannot be estimated directly, more generally where the
forms of statistical dependence among the model's parameters are unknown, making incorrect
assumptions about correlations and dependencies may lead to severe over-estimation 
or under-estimation of extinction risks [@ferson1995correlations]. We believe modeling 
dependencies with copula could help in this case. Second, we did not consider multiple sites of 
age-structured metapopulation which could be linked via dispersal. So, possible future investigation 
would be to explore the viability for the metapopulations if vital rates (e.g. fecundity or 
survival of different stage class) across sites showed specific tail dependence 
with spatially correlated extreme climatic events. Third, we considered tail dependence 
between vital rates when they were positively correlated, but what would be the extinction risk 
and viability of a population if demographic rates were negatively correlated with asymmetric tail dependence. 
This scenario could be considered for those extreme events that kill adults and allow more successful
recruitment of seeds or other propagules. Fourth, following the aforementioned idea, it would be worthy 
to model more complex prey-predator model where asymmetric tail dependent copula could be used to mimic
interspecific-dependence. An earlier study [@carnicer2011widespread] showed extreme events
(i.e., heat waves and droughts) increased mortality and defoliation in southern European forest and its 
dependent herbivorous insects and fungi through trophic linkage. ECEs can even truncate 
multiyear insect outbreak dynamics [@esper20061200]. We believe copula (may be a multidimensional one) has 
a lot more potential to estimate risk of cascading extinction in model food webs.

>>>>

With global climate change, extreme climatic events (ECEs) appear to be more common than usual. These ECEs 
could pose severe impact on the resistance or resilience of a population or community level through
life-history traits of the species. Therefore predicting the impact of ECEs 
on extinction risk of a population is necessery. We here introduced the potential use of "tail-dependence" 
property of a copula in modeling the extinction risk of an age-structured single patch model. We also 
extended the generalized findings of density-independent metapopulation model from @Ghosh_copula 
to density-dependence metapopulation models with this copula approach. For metapopulation models, 
we showed extinction risk not only depended on the nature of tail-dependence of spatially correlated 
noise but was also influenced by the inherent dynamics of the population-model.

@@

<!--Earlier study on metapopulation extinction risk : Large-scale spatial synchrony may also be generated by
other processes such as migration and predation, but regional
stochasticity in the form of spatially correlated weather conditions is
probably the dominant synchronizing mechanism (ref: Nature1998) but what about taildep. effect on extrisk?.-->
In the metapopulation context, regional stochasticity 
can simultaneously affected multiple habitat patches where spatially correlated weather conditions were 
probably the dominant synchronizing mechanism for extinctions [@hanski1998metapopulation]. A 21 years-long
study by @tack2015increasing on the Glanville fritillary butterfly demonstrated that frequency 
of drought during early larval instars increased amplitude and span of spatial synchrony 
over a metapopulation network of 4000 dry meadows and could affect the long-term population viability.
Though earlier literature considered the effect of spatial and/ or temporal correlation in estimating extinction eisk of a metapopulation, but to our knowledge, using the concept of "tail dependence" (i.e. correlation would be stronger during extreme years) to model extinction risk remains unexplored.

<!--How copula could be relevent in estimating extreme event's 
influence on extinction risk : introduce Taildep. concept-->
We recently [@Ghosh_copula] used this tail dependence feature using "copula" - a statistical 
tool broadly used in modeling dependence between variables - to estimate extinction risk of a density-independent 
metapopulation model. Present manuscript is an extension of this concept 
in estimating the extinction risk beyond the density-independent model 
and even applicable for a simple age-structured population where its vital rates are influenced by ECEs. 
Following we will reintroduce the concept of copula in a manner that follows @Ghosh_copula.

<!--brief intro for copula: I am not sure whether it is needed in this paper or not?-->
Copula, a rank based measure of association, is much more informative 
than linear correlation measure and a widely applied tool in dependence 
modeling [@joe2014_dependence; @nelsen2006_copula]. Copulas separate the information content of a bivariate dataset,
$(x_n, y_n)$ for $n = 1 \ldots N$, into two non-overlapping parts: the information in the marginal 
distributions (which is not about the association between the variables) and the rest of 
the information (which is solely about the association). The association information is revealed by 
the copula plot ($v_n$ against $u_n$) of the raw variables ($y_n$ vs. $x_n$) in Fig...., where 
$u_n$ is the rank of $x_n$ in the set ${x_1, x_2, \ldots , x_n }$, divided by $N + 1$; 
and $v_n$ is the rank of $y_n$ in the set ${y_1, y_2, \ldots , y_n }$, also divided
by $N + 1$. Here the rank of the smallest element of each sets is considered to be 1 and the rank of
the largest element is $N$. We refer to the $u_n$ and $v_n$ as normalized ranks of the $x_n$ and $y_n$. 
Ranking makes the marginal distributions uniform, isolating only the information on 
association between the variables.

<!--tail dep. of copula & extinction risk-->
A major benefit of the copula approach is that it can effectively capture the tail dependence. 
Tail dependence, already exemplified in Fig..... (add Fig.), is association between the more extreme 
values of distributions. For two positively associated variables, left-tail 
(respectively, right-tail) dependence is stronger association between
values in the left or lower portions, i. e. for smaller values 
(respectively, the right or upper portions, i.e. for higher values) 
of the two distributions. Stronger dependence in either tail causes asymmetric copula 
structure for the joint distribution of two variables. For example, Clayton and Survival Clayton are 
two copula families with asymmetric tail that are used to model exclusively 
left and right tail dependence between variables, respectively.
From ecological perspective, it is reasonable to expect that extreme cold winter during growing season
might reduce the adult reproduction rate as well as larval survival (i.e. stronger 
tail dependence between vital rates when temperature is too cold than normal) and 
thereby inflate the population viability. Similarly, in context of 
density-dependent metapopulation models, if extreme environmental fluctuations across sites 
appear to be as synchronous (left tail dependent or right tail dependent between patches 
with same correlation as in Fig.....) then it could wipe out some metapopulation patches 
during unfavorable years unless they are rescued by the nearby populations. This scenario 
could be more complex and specific of the direction of dependence 
depending on the dynamics of the model (see \nameref{Results}). Though tail-dependence 
property of a copula has been used for risk-analysis in other fields like finance, 
hydrology etc. [@alexander2008market; @li2013; @chen2011flood; @Goswami2018], but to our 
knowledge asymmetric tail dependence have never been considered 
in estimating the effect of ECEs through extinction risk of age structured 
model or density-dependent metapopulations.

<!--our objective ?-->
In this paper, we will explore the potential use of asymmetric tail-dependence 
for understanding the extinction risk of populations towards extreme climatic event
by addressing the following questions: (Q1) Does nature of tail dependence 
(left tail or right tail) in vital rates
of a single-patch age-structured population model affect extinction risk of that population differently?
We hypothesized population might have higher extinction risk if vital rates are left-tail dependent.
(Q2) How does environmental noises with asymmetric tail dependence across multiple patches
influence extinction risk of a density-dependent metapopulation model? Our speculation 
is that the general finding from density-independent metapopulation model [@Ghosh_copula]
(higher extinction risk if noises across patches were left-tail dependent than the one 
if noises had right-tail dependence) would hold true only in the under-compensatory regime, but 
should be reverse if the model shows over-compensatory dynamics. 
(Q3) We further hypothesize that when there is no dispersal, the above-said switching should occur
near the critical parameter value where the deterministic model shows transition from 
under-compensatory to over-compensatory dynamics.

@@

<!--future study: worthy for PVA, IUCN red list
1. Identify which life stage or demographic process as management target?
2. Old study (Doak1994, Ferguson1995) showed including correlation btw vital rates for PVA makes a difference, 
   what if we include TD through copula? Multisite age str. model with dispersal, directional change
   in spatio-temporal variation for vital rates.
3. Tdep. with -ve correlation between vital rates?
4. extinction risk how propagated through a food web?
5. ref: see McCarthy2000: Most previous models of metapopulation viability have either ignored
correlated extinction of patches, or have included uniform correlations, with the correlation between nearby patches being the same as the correlation between distant
patches (Hanski 1994; Tilman et al. 1994; Lindenmayer et al. 1995; McCarthy
1996). 
6. see this: http://dx.doi.org/10.1098/rspb.2015.0173
-->
Although we demonstrated that tail dependence between vital rates of an age structured model can
influence extinction risk, substantial work remains to determine the importance of this effect in the 
following scenarios. First, we demonstrated the effect of tail-dependence on extinction risk 
with a very simplistic age structured (two age stages) model where we considered the total semelperous 
population would go extinct if there would be only one female adult and one egg. But in reality population 
viability analysis (PVA) deals with a little bit higher threshold with several age, stage or size class and a variety 
of populations including semelperous and iteroperous [@morris2002quantitative]. Future studies should include 
tail-dependence effect on extinction risk 
for a multiple age-, size- and stage- structured model and estimate extinction risk 
for each age, size and stage class separately 
so that we can identify (via multisite-PVA) which life-stage or demographic process 
to target for population management. Earlier study by @doak1994modeling showed PVA models including correlation 
between different demographic rates indicated marked difference in predicting confidence interval for 
future population size than the model which did not include correlation. But when 
such correlation cannot be estimated directly, more generally where the
forms of statistical dependence among the model's parameters are unknown, making incorrect
assumptions about correlations and dependencies may lead to severe over-estimation 
or under-estimation of extinction risks [@ferson1995correlations]. We believe modeling 
dependencies with copula could help in this case. Second, we did not consider multiple sites of 
age-structured metapopulation which could be linked via dispersal. So, possible future investigation 
would be to explore the viability for the metapopulations if vital rates (e.g. fecundity or 
survival of different stage class) across sites showed specific tail dependence 
with spatially correlated extreme climatic events. Third, we considered tail dependence 
between vital rates when they were positively correlated, but what would be the extinction risk 
and viability of a population if demographic rates were negatively correlated with asymmetric tail dependence. 
This scenario could be considered for those extreme events that kill adults and allow more successful
recruitment of seeds or other propagules. Fourth, following the aforementioned idea, it would be worthy 
to model more complex prey-predator model where asymmetric tail dependent copula could be used to mimic
interspecific-dependence. An earlier study [@carnicer2011widespread] showed extreme events
(i.e., heat waves and droughts) increased mortality and defoliation in southern European forest and its 
dependent herbivorous insects and fungi through trophic linkage. ECEs can even truncate 
multiyear insect outbreak dynamics [@esper20061200]. We believe copula (may be a multidimensional one) has 
a lot more potential to estimate risk of cascading extinction in model food webs.
